An antibody-escape estimator for mutations to the SARS-CoV-2 receptor-binding domain

### 2022-03-07 paper/reviewer_response.tex

The most important aspect of the paper is the correlation between the calculated binding score and the experimentally-measure fold change in virus neutralization as presented in Figure 3, which would provide some kind of validation of the computational method.  The authors state in the text that “the escape-calculator scores correlate quite well with the measured neutralization titers across all studies and cohorts”; “quite well” seems pretty subjective.  At a minimum, the strength of correlation should be quantified by using something like a Spearman’s rho metric which could then be used to calculate a t test statistic and p-value.  Indeed, these metrics could be used to compare the performance of the different configuration parameters.

\response{
We have added Pearson's R and the associated P-values in Figure 3.
Five of the 6 correlations are significant, with $P < 0.05$.
In addition, we have used correlation coefficients to compare the sum-of-mutations-at-a-site model to the mean-of-mutations-at-a-site model as described in the point immediately above.
}

It is not clear how to use the online calculator.  The instructions online refer to a Tweet chain but that provides a summary of how the calculations are performed, not on how to use the online tool.  It is not very intuitive and could use some simple explicit instructions on the page itself.

\response{We have added more instructions on the online page.}

@@

Greaney et al have developed a simple tool for estimating the impact of amino acid changes on antibody protection from SAR-CoV-2 variants. This is based on experimental data, much of which has been generated by these researchers. The text is clear and well written, the method is both sound and an easy to use implementation is provided. I only have some minor comments:

\response{Thanks for the positive assessment.}

Please make it clear in the abstract that the experimental data is generated on the early pre-variant of concern data and so mutations may differ in consequence on new variants such as Omicron. This would seem to be quite a big caveat with this type of data.

\response{This is a good point, and we have added the following sentence to the Abstract: \emph{Currently the calculator uses primarily data for antibodies elicited by Wuhan-Hu-1-like vaccination or infection, and so is expected to work best for calculating escape from such immunity for mutations relative to early SARS-CoV-2 strains.}. We do note that we anticipate updating as new data become available.}

@@

Line 30, variants with reduced Ab neutralisation properties emerged before 2 years, e.g, Beta was designated in late 2020 in South Africa.

\response{Good point, we have clarified to indicate that new variants emerged within a year.}

@@

Line 41, briefly define variants of concern as there's ongoing confusion between these and just variants.

\response{We have clarified that variants of concern are variants with reduced immune recognition, enhanced transmissibility, or increased virulence.}

@@

Line 54, the pop here at "complex black-box computational methods" is unwarranted and out of date relative to much contemporary machine leaning methods. Arguably simple residue based information from experiments is also a black-box as the mechanistic nuance of three-dimensional structural interactions are not really being accounted for. Maybe the authors could comment in the Discussion on why such a simple method works so well, e.g., linked to immunodominance of this spike region.

\response{We have removed the word ``black box'' and clarified that we just mean that the computations can be interpreted visually using the online calculator. We have also elaborated on immunodominance.}

@@

Two lines into Results, mutations strictly speaking occur at the nucleotide level and result in amino acid replacements. Clearly the authors know this and it fine to use mutations as short-hand for the residue changes but at least make the distinction clear the first time this is referred to.

\response{We have clarified this directly: \emph{... where we here consider mutations at the level of amino-acid replacements rather than in terms of the underlying nucleotide changes.}}

@@

Results, 3rd and 4th paragraph, let's not 'imagine' anything, please re-write.

\response{We have re-worded to make clear that we mean to consider a hypothetical antibody mix.}

@@

Page 2, "illustrative" would be more accurate than "toy" example.

\response{We have changed ``toy'' to ``illustrative'' as suggested.}

### 2022-03-07 paper/paper.tex
responses to reviewer 2

\begin{abstract}
A key goal of SARS-CoV-2 surveillance is to rapidly identify viral variants with mutations that reduce neutralization by polyclonal antibodies elicited by vaccination or infection.
Unfortunately, direct experimental characterization of new viral variants lags their sequence-based identification. 
Here we help address this challenge by aggregating deep mutational scanning data into an ``escape estimator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
The estimator can be used to intuitively visualize how mutations impact polyclonal antibody recognition, and score the expected antigenic effect of combinations of mutations.
These scores correlate with neutralization assays performed on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
An interactive version of the estimator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
\end{abstract}

@@

\lettrine[lines=2]{\color{color2}H}{}uman coronaviruses undergo antigenic evolution that erodes antibody-based neutralization~\citep{eguia2021human,kistler2021evidence}.
This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization have emerged only $\sim$2 years since the virus first started to spread in humans.
A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{wang2021antibody,uriu2021neutralization,lucas2021impact}.
Unfortunately, the rate at which new variants arise outstrips the speed at which these experiments can be performed.

A partial solution is to use deep mutational scanning experiments to \emph{prospectively} measure how viral mutations impact antibody binding or neutralization.
Deep mutational scanning can systematically measure the antigenic impacts of all possible amino-acid mutations in key regions of spike on monoclonal antibodies~\citep{starr2021prospective,greaney2021complete} or sera~\citep{greaney2021comprehensive}.
However, SARS-CoV-2 variants of concern typically have multiple mutations, and it is not feasible to experimentally characterize all combinations of mutations even via high-throughput approaches such as deep mutational scanning.

Here we take a step towards addressing this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high}.
The resulting ``escape estimator'' enables qualitative visualization and quantitative scoring of the antigenic effects of arbitrary combinations of mutations.
Importantly, the escape estimator is based on simple transformations of direct experimental measurements, and does not involve complex black-box computational methods.

@@

\subsection{Combining monoclonal antibody escape maps reveals correlated and independent viral antigenic mutations}
A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody~\citep{greaney2021complete}.
This mutation-level information can be summarized for each RBD site, such as by taking the mean or sum of mutation-level effects at a site.
Here we will work with these site-level escape maps.
We use site-level information for two reasons: mutation-level measurements tend to be more noisy and averaging them for a site decreases this noise, and using site-level information makes the approach independent of the particular wildtype amino acid at a site (which is useful if we want to keep using the calculator as the RBD evolves).
However, we note that site-level information ignores the possibility of different mutations at a site having different effects, and so mutation-level approaches could also become useful as the quality of experimental data improve.

@@

Now imagine a polyclonal antibody mix of these three antibodies combined at equal potencies.
We can generate an escape map for this hypothetical antibody mix simply by averaging the experimentally measured escape maps for the three individual antibodies, yielding the thick black line in Figure~\ref{fig:mini_example}A.
Because this polyclonal escape map is the average of the monoclonal antibody maps, its largest peaks are at the sites of strongest escape for each individual antibody: 417, 484, and 444--446.

Next imagine removing one antibody from the mix by mutating its epitope.
Figure~\ref{fig:mini_example}B shows the resulting escape map if LY-CoV555 is ablated, as would occur if site 484 was mutated.
The thick black line for the antibody mix no longer has peaks at 484 and other sites targeted by LY-CoV555, such as 490.
Therefore, in this hypothetical polyclonal antibody mix, escape at sites 484 and 490 is correlated since both sites are targeted by the same antibody.
However, the polyclonal mix's escape map at sites 417 and 460 is unaffected by mutations that escape LY-CoV555, since they are targeted by a different antibody, LY-CoV016.
But if we also ablate LY-CoV016 (such as by mutating site 417), then the peaks at 417 and 460 also disappear, and the remaining peaks are at sites targeted by REGN10987, such as 444--446 (Figure~\ref{fig:mini_example}C).
Of course, if REGN10987 was also ablated such as by mutating site 446, then the polyclonal antibody mix would have no remaining activity.
This and other scenarios can be explored using the interactive version of Figure~\ref{fig:mini_example} at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.

@@

\subsection{Aggregating deep mutational scanning data for 33 human antibodies yields a realistic escape estimator}
The toy example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
To create an escape map for an antibody mix that more realistically represents actual human sera, we aggregated previously generated deep mutational scanning data for 33 neutralizing antibodies elicited by SARS-CoV-2.
These antibodies were isolated from a variety of patient cohorts within the first year of the pandemic (see Methods for details).
An assumption of the analysis that follows is that an equipotent mixture of these 33 antibodies represents the neutralizing activity of human sera; we emphasize that this assumption is imperfect since in reality the antibodies were chosen for prior study for a variety of ad hoc reasons.
The escape maps for all the individual antibodies can be interactively interrogated at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/}.

@@

The overall polyclonal escape map generated by averaging the experimental data for all 33 antibodies is in Figure \ref{fig:escape_calc}A.
As in the toy three-antibody example in the previous section, there are peaks at  sites 417, 484, and 444--446.
However, the peak at 484 is now larger than any other peak, reflecting the fact that antibodies targeting the class 2 epitope containing E484 are especially common in the human antibody response to early SARS-CoV-2 strains~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.
In addition, there are smaller peaks at a variety of other sites, reflecting the fact that each antibody has a somewhat idiosyncratic epitope (Figure \ref{fig:escape_calc}A).

We can follow the principle outlined in the toy example in the previous section to calculate the expected polyclonal escape map \emph{after} mutating sites in the RBD.
Specifically, we reduce the contribution of each antibody by an amount that scales with how strongly that antibody targets each mutated site (see Methods for details).
For instance, the blue lines in Figure~\ref{fig:escape_calc}B show the polyclonal escape map after mutating site 484.
Mutating site 484 obviously drops the contribution of that site, but it also decreases the contribution of other sites such as 490 that are commonly targeted by antibodies with epitopes that include site 484.
In contrast, mutating site 484 has minimal effect on the polyclonal escape map at sites like 417 or 444--446, since those sites are generally targeted by antibodies that are unaffected by mutations at site 484.

@@

Another key aspect of our approach is the interactive visual implementation of the escape estimator, so the user can interrogate the effects of mutations (or combinations of mutations) simply by clicking (or shift-clicking) on sites.
This interactivity provides an intuitive understanding of the antigenic structure of the RBD, and shows how the estimator is performing simple transformations directly on experimental data.
We encourage interactive use of the estimator so it acts as an aid to augment human interpretation, rather than a black-box computational method.
However, we note that in addition to the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}.

>>>>

\begin{abstract}
A key goal of SARS-CoV-2 surveillance is to rapidly identify viral variants with mutations that reduce neutralization by polyclonal antibodies elicited by vaccination or infection.
Unfortunately, direct experimental characterization of new viral variants lags their sequence-based identification. 
Here we help address this challenge by aggregating deep mutational scanning data into an ``escape estimator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
The estimator can be used to intuitively visualize how mutations impact polyclonal antibody recognition, and score the expected antigenic effect of combinations of mutations.
These scores correlate with neutralization assays performed on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
An interactive version of the estimator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
Currently the calculator uses primarily data for antibodies elicited by Wuhan-Hu-1-like vaccination or infection, and so is expected to work best for calculating escape from such immunity for mutations relative to early SARS-CoV-2 strains.
\end{abstract}

@@

\lettrine[lines=2]{\color{color2}H}{}uman coronaviruses undergo antigenic evolution that erodes antibody-based neutralization~\citep{eguia2021human,kistler2021evidence}.
This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization emerged within a year of when the virus first started to spread in humans.
A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{wang2021antibody,uriu2021neutralization,lucas2021impact}.
Unfortunately, the rate at which new variants arise outstrips the speed at which these experiments can be performed.

A partial solution is to use deep mutational scanning experiments to \emph{prospectively} measure how viral mutations impact antibody binding or neutralization.
Deep mutational scanning can systematically measure the antigenic impacts of all possible amino-acid mutations in key regions of spike on monoclonal antibodies~\citep{starr2021prospective,greaney2021complete} or sera~\citep{greaney2021comprehensive}.
However, SARS-CoV-2 variants of concern (variants with reduced immune recognition, enhanced transmissibility, or increased virulence) typically have multiple mutations, and it is not feasible to experimentally characterize all combinations of mutations even via high-throughput approaches such as deep mutational scanning.

Here we take a step towards addressing this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high}.
The resulting ``escape estimator'' enables qualitative visualization and quantitative scoring of the antigenic effects of arbitrary combinations of mutations.
Importantly, the escape estimator is based on simple transformations of direct experimental measurements, and and so its calculations can be intuitively visualized using the interactive interface we provide.

@@

\subsection{Combining monoclonal antibody escape maps reveals correlated and independent viral antigenic mutations}
A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody~\citep{greaney2021complete}, where we here consider mutations at the level of amino-acid replacements rather than in terms of the underlying nucleotide changes.
This mutation-level information can be summarized for each RBD site, such as by taking the mean or sum of mutation-level effects at a site.
Here we will work with these site-level escape maps.
We use site-level information for two reasons: mutation-level measurements tend to be more noisy and averaging them for a site decreases this noise, and using site-level information makes the approach independent of the particular wildtype amino acid at a site (which is useful if we want to keep using the calculator as the RBD evolves).
However, we note that site-level information ignores the possibility of different mutations at a site having different effects, and so mutation-level approaches could also become useful as the quality of experimental data improve.

@@

Now consider a hypothetical polyclonal antibody mix of these three antibodies combined at equal potencies.
We can generate an escape map for this hypothetical antibody mix simply by averaging the experimentally measured escape maps for the three individual antibodies, yielding the thick black line in Figure~\ref{fig:mini_example}A.
Because this polyclonal escape map is the average of the monoclonal antibody maps, its largest peaks are at the sites of strongest escape for each individual antibody: 417, 484, and 444--446.

Next consider removing one antibody from the hypothetical mix by mutating its epitope.
Figure~\ref{fig:mini_example}B shows the resulting escape map if LY-CoV555 is ablated, as would occur if site 484 was mutated.
The thick black line for the antibody mix no longer has peaks at 484 and other sites targeted by LY-CoV555, such as 490.
Therefore, in this hypothetical polyclonal antibody mix, escape at sites 484 and 490 is correlated since both sites are targeted by the same antibody.
However, the polyclonal mix's escape map at sites 417 and 460 is unaffected by mutations that escape LY-CoV555, since they are targeted by a different antibody, LY-CoV016.
But if we also ablate LY-CoV016 (such as by mutating site 417), then the peaks at 417 and 460 also disappear, and the remaining peaks are at sites targeted by REGN10987, such as 444--446 (Figure~\ref{fig:mini_example}C).
Of course, if REGN10987 was also ablated such as by mutating site 446, then the polyclonal antibody mix would have no remaining activity.
This and other scenarios can be explored using the interactive version of Figure~\ref{fig:mini_example} at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.

@@

\subsection{Aggregating deep mutational scanning data for 33 human antibodies yields a realistic escape estimator}
The illustrative example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
To create an escape map for an antibody mix that more realistically represents actual human sera, we aggregated previously generated deep mutational scanning data for 33 neutralizing antibodies elicited by SARS-CoV-2.
These antibodies were isolated from a variety of patient cohorts within the first year of the pandemic (see Methods for details).
An assumption of the analysis that follows is that an equipotent mixture of these 33 antibodies represents the neutralizing activity of human sera; we emphasize that this assumption is imperfect since in reality the antibodies were chosen for prior study for a variety of ad hoc reasons.
The escape maps for all the individual antibodies can be interactively interrogated at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/}.

@@

The overall polyclonal escape map generated by averaging the experimental data for all 33 antibodies is in Figure \ref{fig:escape_calc}A.
As in the illustrative three-antibody example in the previous section, there are peaks at  sites 417, 484, and 444--446.
However, the peak at 484 is now larger than any other peak, reflecting the fact that antibodies targeting the class 2 epitope containing E484 are especially common in the human antibody response to early SARS-CoV-2 strains~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.
In addition, there are smaller peaks at a variety of other sites, reflecting the fact that each antibody has a somewhat idiosyncratic epitope (Figure \ref{fig:escape_calc}A).

We can follow the principle outlined in the illustrative example in the previous section to calculate the expected polyclonal escape map \emph{after} mutating sites in the RBD.
Specifically, we reduce the contribution of each antibody by an amount that scales with how strongly that antibody targets each mutated site (see Methods for details).
For instance, the blue lines in Figure~\ref{fig:escape_calc}B show the polyclonal escape map after mutating site 484.
Mutating site 484 obviously drops the contribution of that site, but it also decreases the contribution of other sites such as 490 that are commonly targeted by antibodies with epitopes that include site 484.
In contrast, mutating site 484 has minimal effect on the polyclonal escape map at sites like 417 or 444--446, since those sites are generally targeted by antibodies that are unaffected by mutations at site 484.

@@

Another key aspect of our approach is the interactive visual implementation of the escape estimator, so the user can interrogate the effects of mutations (or combinations of mutations) simply by clicking (or shift-clicking) on sites.
This interactivity provides an intuitive understanding of the antigenic structure of the RBD, and shows how the estimator is performing simple transformations directly on experimental data.
We encourage interactive use of the estimator so it acts as a visual aid to augment human interpretation.
However, we note that in addition to the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}.
